---
title: "R Graphics - High-Density Scatterplots"
subtitle: "Solutions for Large Datasets and Overplotting"
author: "Stephen Moore"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    html_document:
        theme: default
        highlight: tango
        toc: true        
---

### Introduction

This document demonstrates different ways of generating scatter plots for large datasets with the ggplot2 and tabplot plotting packages.  Plots make use of the diamonds dataset.

```{r init, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(tidy=T, 
               fig.width=10,
               fig.height=5,
               fig.align='left',
               warning=FALSE,
               message=FALSE,
               echo=TRUE)
options(width = 120)
library(ggplot2)
library(colorspace)
library(gridExtra)
attach(diamonds)
```

## The Overplotting Problem

Standard scatterplots of large datasets are blobs that obscure structure in the data. We can mitigate the problem somewhat by manipulating the alpha transparency of geom_point().

```{r overplotting1, tidy=F}
p = ggplot(diamonds,aes(x=carat,y=price))  +
  ggtitle("Price vs. Carat Weight") +
  xlab("Carat Weight") +
  ylab("Price")
p1 = p + geom_point()
p2 = p + geom_point(alpha = 0.1, colour="purple") +
  theme_bw()
grid.arrange(p1,p2,ncol=2)
```

## Better Solutions to Overplotting

### ggplot2 Scatterplots

Using density lines or heatmaps can reveal structure in large datasets.  In addition, a log10 transform of the price and carat variables also helps to reveal clusters and price points around standard carat weights (1/4, 1/2, 1, 1.5, 2, 3).

```{r overplotting2, tidy=F}
p = ggplot(diamonds,aes(x=log10(carat),y=log10(price))) +
  ggtitle("Price vs. Carat Weight") +
  xlab("log10(Carat Weight)") +
  ylab("log10(Price)") +
  ylim(2.5,4.5)
p1 = p + 
  geom_point(alpha = 0.1, colour="orange") + 
  geom_density2d()+ theme_bw()
p2 = p +
  geom_density2d()+ theme_bw()
grid.arrange(p1,p2,ncol=2)

p1 = p +
  stat_binhex(colour="white", na.rm=TRUE) +
  scale_fill_gradientn(colours=c("purple","green"), name = "Frequency", na.value=NA)
p2 = p +
  stat_binhex(colour="white", na.rm=TRUE) +
  scale_fill_gradientn(colours=c("blue","orange"), name = "Frequency", na.value=NA) + 
  theme_bw()
grid.arrange(p1,p2,ncol=2)

p1 = ggplot(diamonds,aes(x=log10(carat),y=log10(price),colour=color)) +
  ggtitle("Price vs. Carat Weight") +
  xlab("log10(Carat Weight)") +
  ylab("log10(Price)") +
  ylim(2.5,4.5) + 
  geom_point()
p2 = ggplot(diamonds,aes(x=log10(carat),y=log10(price),colour=cut)) +
  ggtitle("Price vs. Carat Weight") +
  xlab("log10(Carat Weight)") +
  ylab("log10(Price)") +
  ylim(2.5,4.5) + 
  geom_point()
grid.arrange(p1,p2,ncol=2)
```

### Smoothed Scatterplots

```{r smoothedScatterplot, fig.width=6, tidy=F}

# create a color palette to use in smoothed scatterplot
library(RColorBrewer)
buylrd = c("#313695", "#4575B4", "#74ADD1", "#ABD9E9", "#E0F3F8", "#FFFFBF",
           "#FEE090", "#FDAE61", "#F46D43", "#D73027", "#A50026") 
myColRamp = colorRampPalette(c(buylrd))

# smoothed scatterplot
smoothScatter(x=log10(carat),y=log10(price),
              colramp=myColRamp,
              main="Price vs. Caret Weight",
              xlab="log10(Carat Weight)",
              ylab="log10(Price)")
```

### Visualization of large datasets with tabplot

```{r tabplot, tidy=F}
library(tabplot)
tableplot(diamonds, select = c(carat, price, cut, color, clarity), sortCol = price)
```

### Programming Environment

```{r}
sessionInfo()
```


