---
title: "R Graphics - High-Density Scatterplots"
subtitle: "Solutions for Large Datasets and Overplotting"
author: "Stephen Moore"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    html_document:
        theme: default
        highlight: tango
        toc: true        
---

### Introduction

This document demonstrates different ways of generating scatter plots for large datasets with the ggplot2 and tabplot plotting packages.  Plots make use of the diamonds dataset.

```{r init, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(tidy=FALSE, 
               fig.width=10,
               fig.height=5,
               fig.align='left',
               warning=FALSE,
               message=FALSE,
               echo=TRUE)
options(width = 120)
library(ggplot2)
library(colorspace)
library(gridExtra)
# Note:  package hexbin must be installed for hex plot below to work
```

## The Overplotting Problem

Standard scatterplots of large datasets are blobs that obscure structure in the data. We can mitigate the problem somewhat by manipulating the alpha transparency of geom_point(), which begins to reveal the varying density of data points
.
```{r, cache=TRUE}
# generate an artificial, large dataset
x1 = rnorm(100000)
x2 = rnorm(100000)
bigdf = as.data.frame(cbind(x1,x2))
```

```{r overplotting1}
p = ggplot(bigdf,aes(x=x1,y=x2))  +
  ggtitle("Plot of 100K Point Dataset") +
  xlab("x1") +
  ylab("x2")
p1 = p + geom_point()
p2 = p + geom_point(alpha = 0.01, colour="purple") +
  theme_bw()
grid.arrange(p1,p2,ncol=2)
```

## Better Solutions to Overplotting

### ggplot2 Scatterplots

Using density lines or heatmaps can reveal structure in large datasets.

```{r overplotting2}
p = ggplot(bigdf,aes(x=x1,y=x2)) +
  ggtitle("Plot of 100K Point Dataset") +
  xlab("x1") +
  ylab("x2")
p1 = p + 
  geom_point(alpha = 0.01, colour="orange") + 
  geom_density2d() + 
  theme_bw()
p2 = p +
  # geom_hex() simple
  geom_hex(colour="white", na.rm=TRUE) +
  scale_fill_gradientn(colours=c("purple","green"),
                       name = "Frequency",
                       na.value=NA)
grid.arrange(p1,p2,ncol=2)
```

### Smoothed Scatterplots

```{r smoothedScatterplot, fig.width=6, fig.height=6}

# create a color palette to use in smoothed scatterplot
library(RColorBrewer)
buylrd = c("#313695", "#4575B4", "#74ADD1", "#ABD9E9", "#E0F3F8", "#FFFFBF",
           "#FEE090", "#FDAE61", "#F46D43", "#D73027", "#A50026") 
myColRamp = colorRampPalette(c(buylrd))

# smoothed scatterplot
smoothScatter(x=x1,y=x2,
              colramp=myColRamp,
              main="Plot of 100K Point Dataset",
              xlab="x1",
              ylab="x2")
```

## Programming Environment

```{r}
sessionInfo()
```


